---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, results='markup', warning=TRUE, message=TRUE}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
activityData <- read.csv('activity.csv')
```
```{r}
#activityData$interval <- strptime(gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", activityData$interval), format='%H:%M')
```
```{r}


## What is mean total number of steps taken per day?


mean_steps <- mean(activity_steps_day$steps)
median_steps <- median(activity_steps_day$steps)
#we set a normal number format to display the results
mean_steps <- format(mean_steps,digits=1)
median_steps <- format(median_steps,digits=1)



## What is the average daily activity pattern?
```{r}
stepsByDay <- tapply(activityData$steps, activityData$date, sum, na.rm=TRUE)
```



## Imputing missing values
sum(is.na(activity))
#subset general dataset with missing values only
missing_values <- subset(activity, is.na(steps))
#plot repartition, by date or by intervals
par(mfrow = c(2,1), mar = c(2, 2, 1, 1))
hist(missing_values$interval, main="NAs repartition per interval")
hist(as.numeric(missing_values$date), main = "NAs repartion per date", breaks = 61)
# calculate mean of steps per interval, we end up with a mean for all 288 intervals
MeanStepsPerInterval <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
# cut the 'activity' dataset in 2 parts (with and without NAs)
activity_NAs <- activity[is.na(activity$steps),]
activity_non_NAs <- activity[!is.na(activity$steps),]
# replace missing values in activity_NAs
activity_NAs$steps <- as.factor(activity_NAs$interval)
levels(activity_NAs$steps) <- MeanStepsPerInterval
# change the vector back as integer 
levels(activity_NAs$steps) <- round(as.numeric(levels(activity_NAs$steps)))
activity_NAs$steps <- as.integer(as.vector(activity_NAs$steps))
# merge/rbind the two datasets together
imputed_activity <- rbind(activity_NAs, activity_non_NAs)
# Plotting parameters to place previous histogram and new one next to each other
par(mfrow = c(1,2))
# Plot again the histogram from the first part of the assignment
activity_steps_day <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
hist(activity_steps_day$steps, xlab = "Steps per Day", main = "NAs REMOVED - Total steps/day", col = "wheat")
# Plot new histogram, with imputed missing values
imp_activity_steps_day <- aggregate(steps ~ date, data = imputed_activity, FUN = sum, na.rm = TRUE)
hist(imp_activity_steps_day$steps, xlab = "Steps per Day", main = "NAs IMPUTED - Total steps/day", col = "wheat")
imp_mean_steps <- mean(imp_activity_steps_day$steps)
imp_median_steps <- median(imp_activity_steps_day$steps)
# we set a normal number format to display the results
imp_mean_steps <- format(imp_mean_steps,digits=1)
imp_median_steps <- format(imp_median_steps,digits=1)
# store the results in a dataframe
results_mean_median <- data.frame(c(mean_steps, median_steps), c(imp_mean_steps, imp_median_steps))
colnames(results_mean_median) <- c("NA removed", "Imputed NA values")
rownames(results_mean_median) <- c("mean", "median")
library(xtable)
xt <- xtable(results_mean_median)
print(xt, type  = "html")


## Are there differences in activity patterns between weekdays and weekends?
#elseif function to categorize Saturday and Sunday as factor level "weekend", all the rest as "weekday"
imputed_activity$dayType <- ifelse(weekdays(as.Date(imputed_activity$date)) == "Samstag" | weekdays(as.Date(imputed_activity$date)) == "Sonntag", "weekend", "weekday")
#transform dayType variable into factor
imputed_activity$dayType <- factor(imputed_activity$dayType)
#Aggregate a table showing mean steps for all intervals, acrlss week days and weekend days
steps_interval_dayType <- aggregate(steps ~ interval + dayType, data = imputed_activity, FUN = mean)
#verify new dataframe 
head(steps_interval_dayType)
#add descriptive variables
names(steps_interval_dayType) <- c("interval", "day_type", "mean_steps")
#plot with ggplot2
library(ggplot2)
plot <- ggplot(steps_interval_dayType, aes(interval, mean_steps))
plot + geom_line(color = "tan3") + facet_grid(day_type~.) + labs(x = "Intervals", y = "Average Steps", title = "Activity Patterns")
